Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-17 paper.qmd

XXX TO DO:

- Remove all the extraneous commented text (NOT Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible
- Check spelling is consistently American
- Check references
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

```{r}
#| label: tbl-notation
#| tbl-cap: "Summary of notation for describing new methodology."
# Notation used in the paper

notation_df <- read_csv("misc/notation.csv")

# Create the table
kable(notation_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("3cm", "12cm"))
```

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

## First projection
model_prj1 <- cbind(
  c(0.37847,-0.06517,0.04231,0.01655),
  c(-0.06466,-0.27052,0.22131,-0.15235))

proj_obj1 <- quollr::get_projection(projection = model_prj1, 
               proj_scale = 1, 
               highd_data = scaled_two_curvy_data, 
               model_highd = scaled_two_curvy_data_model, 
               trimesh_data = tr_from_to_df_two_curvy, 
               axis_param = list(limits = 0.5, 
                                 axis_scaled = 2,
                                 axis_pos_x = -0.5, 
                                 axis_pos_y = -0.5, 
                                 threshold = 0.03))

## For true model

projection_scaled <- model_prj1 * 1

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% projection_scaled

projected_true_model_df <- projected_true_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

true_model_df_proj <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_model_df, 
  by = c("from" = "ID"))

names(true_model_df_proj)[3:NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_from")

true_model_df_proj <- dplyr::left_join(true_model_df_proj, projected_true_model_df, by = c("to" = "ID"))
names(true_model_df_proj)[(2 + NCOL(projected_true_model_df)):NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_to")

proj_obj1_true <- proj_obj1
proj_obj1_true[["model_df"]] <- true_model_df_proj

two_curvy_proj_true <- plot_proj(
  proj_obj = proj_obj1_true, 
  point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  axis_text_size = 4, 
  is_category = FALSE) +
  interior_annotation(label = "a", cex = 1.5) +
  theme(aspect.ratio=1, 
        legend.position = "none")

# Changed the axis parameters
axis_obj <- gen_axes(
    proj = model_prj1 * 2,
    limits = 0.9,
    axis_pos_x = -0.4,
    axis_pos_y = -0.4,
    axis_labels = names(scaled_two_curvy_data),
    threshold = 0.05)

axes <- axis_obj$axes
circle <- axis_obj$circle

proj_obj1[["axes"]] <- axes
proj_obj1[["circle"]] <- circle

two_curvy_proj_tsne_first_model1 <- plot_proj(
    proj_obj = proj_obj1, 
    point_param = c(0.5, 0.2, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    axis_text_size = 2, 
    is_category = FALSE) +
  interior_annotation(label = "b", position = sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1,
        legend.position = "none")

## Second projection
model_prj2 <- cbind(
  c(-0.38281,-0.04870,0.00619,-0.02451),
  c(-0.01303,0.00634,-0.37421,0.09648))

proj_obj2 <- get_projection(projection = model_prj2, 
                            proj_scale = 1, 
                            highd_data = scaled_two_curvy_data, 
                            model_highd = scaled_two_curvy_data_model, 
                            trimesh_data = tr_from_to_df_two_curvy, 
                            axis_param = list(limits = 0.35, 
                                              axis_scaled = 2,
                                              axis_pos_x = -0.35, 
                                              axis_pos_y = -0.35, 
                                              threshold = 0.01))

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% model_prj2

projected_true_model_df <- projected_true_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

true_model_df_proj <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_model_df, 
  by = c("from" = "ID"))

names(true_model_df_proj)[3:NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_from")

true_model_df_proj <- dplyr::left_join(true_model_df_proj, projected_true_model_df, by = c("to" = "ID"))
names(true_model_df_proj)[(2 + NCOL(projected_true_model_df)):NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_to")

proj_obj2_true <- proj_obj2
proj_obj2_true[["model_df"]] <- true_model_df_proj

two_curvy_proj_true2 <- plot_proj(
    proj_obj = proj_obj2_true, 
    point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "black"), # linewidth, alpha
    plot_limits = c(-0.42, 0.42), 
    axis_text_size = 4, 
    is_category = FALSE) +
  interior_annotation(label = "b", cex = 1.5) +
  theme(aspect.ratio = 1, 
        legend.position = "none")

#Changed the axis parameters
axis_obj <- gen_axes(
    proj = model_prj2 * 2,
    limits = 0.7,
    axis_pos_x = -0.35,
    axis_pos_y = -0.35,
    axis_labels = names(scaled_two_curvy_data),
    threshold = 0.02)

axes <- axis_obj$axes
circle <- axis_obj$circle

proj_obj2[["axes"]] <- axes
proj_obj2[["circle"]] <- circle

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", position = sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1) #+
  #geom_segment(
  #  data = model_df, 
  #  aes(
  #    x = proj1_from, 
  #    y = proj2_from, 
  #    xend = proj1_to, 
  #    yend = proj2_to), 
  #  color = "#000000")

@@

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two \\gD{} projections from $4\\text{-}D$, for the two non-linear clusters data (blue points) and true model (black lines) are shown. The top cluster in (a) shows uniformly distributed data along a C-shape, while in the bottom cluster, the data is more concentrated in one corner of the curvy shape. Video of the langevitour animations are available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5
#| eval: false

two_curvy_proj_true + two_curvy_proj_true2 +
  plot_layout(ncol = 2)
```

@@

```{r}
#| label: error-two-curvy

error_two_curvy_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")

## Find the minimum MSE when have duplicate a1
error_two_curvy_umap <- error_two_curvy_umap |>
  group_by(a1) |>
  filter(bin1 == min(bin1)) |>
  ungroup() |>
  mutate(prop_dens = 1/(b*side_length^2))

base_line_prop_dens <- error_two_curvy_umap |>
  filter(a1 == min(a1)) |>
  pull(prop_dens)

error_two_curvy_umap <- error_two_curvy_umap |>
  mutate(prop_comp = prop_dens/base_line_prop_dens)

@@

A new benefit of this fitted model is that it allows us to now predict a new observation's value in the NLDR, for any method. The steps are to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$. This can be written as, let $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$, then the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.

<!--prediction for true-model of two-curvy clust-->
```{r}
#| label: model-prediction-umap-original

predict_tsne_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_umap_predict_true.rds")

predict_two_curvy_obj <- gen_scaled_data(
    nldr_data = predict_tsne_two_curvy)

predict_tsne_two_curvy_scaled <- predict_two_curvy_obj$scaled_nldr

plot_predict_umap_fun_two_curvy <- ggplot(
  data = predict_tsne_two_curvy_scaled, 
  aes(x = UMAP1, 
      y = UMAP2)) +
  geom_point(alpha = 0.5) +
  interior_annotation("b", c(0.08, 0.93), cex = 1.5)
```

```{r}
#| label: model-prediction-points-line
#| eval: false

predict_tsne_two_curvy_scaled <- predict_tsne_two_curvy_scaled |>
  mutate(ID = row_number()) |>
  mutate(type = "predict_from_umap")

jittered_points_df <- jittered_points_df |>
  mutate(type = "predict_from_model")

predict_point_df <- bind_rows(
  predict_tsne_two_curvy_scaled,
  jittered_points_df
)

plot_predict_umap_connect_two_curvy <- ggplot(
  data = predict_point_df, 
  aes(x = UMAP1, 
      y = UMAP2, 
      group = ID)) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.3) +
  interior_annotation("d", c(0.08, 0.93), cex = 1.5)
```

```{r}
#| echo: false
#| label: fig-predict-two-curvy
#| fig-pos: H
#| fig-width: 10
#| fig-height: 10
#| out-width: 80%
#| fig-cap: "Comparison of prediction generated using the exiting `umap` R package's prediction method and our method: (a) A view of the true model in projection from $7\\text{-}D$, (b) predicted data from the `umap` prediction method, and (c) predicted data from our method."
#| eval: false

proj_plot_two_curvy_true + plot_predict_umap_fun_two_curvy +plot_predict_tsne_two_curvy +
  plot_layout(ncol = 3)
```

@@

>>>>

<!-- 
XXX TO DO:

- Remove all the extraneous commented text (NOT Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible
- Check spelling is consistently American
- Check references
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.
-->

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)~parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

## First projection
proj_obj1 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj1.rds")

two_curvy_proj_tsne_first_model1 <- plot_proj(
    proj_obj = proj_obj1, 
    point_param = c(0.5, 0.2, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.6, 0.6), 
    axis_text_size = 2, 
    is_category = FALSE) +
  interior_annotation(label = "b", cex = 1.2) +
  theme(aspect.ratio = 1,
        legend.position = "none")

## Second projection

proj_obj2 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj2.rds")

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", cex = 1.2) +
  theme(aspect.ratio = 1)

@@

@@

```{r}
#| label: error-two-curvy
error_two_curvy_umap <- read_rds("data/two_nonlinear/error_two_curvy_umap.rds")

@@

<!-- Does this really need an entire subsection? It's only 3 sentences long. -->

In this paradigm of model fitting, it is now possible to predict a new observation's value in the NLDR representation, independent of the choice of NLDR method. To do this, we first need to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$, to the new observation, which is,

$$z = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}.$$

Then, the predicted value is is the centroid of this bin in \gD{}, $C^{(2)}_{z} = (c_{z1}, c_{z2})$.

@@

Direct links to videos for viewing online are available in @tbl-links.

```{r}
#| label: tbl-links
#| tbl-cap: "Videos of the langevitour animations and the linked plots."
# Video links in the paper

links_df <- read_csv("misc/video_links.csv")

# Create the table
kable(links_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:3, width = c("1.5cm", "4.5cm", "9cm"))
```
